---
title: "Reproducible Research Project 1"
author: "Neil"
output: html_document
---


# Project Outline:
Using randomised data from a personal activity monitoring device, the supplied dataset records data a 5 minute intervals throughout the day. The data consists of two months of data, from Oct & Nov 2012, & includes the number of steps taken in 5 minute intervals. 

This assignment covers the following questions:

1. Load & process the data
2. What is the mean total number of steps taken per day?
3. What is the average daily activity pattern?
4. Imputing missing values
5. Are there differences in activity patterns between weekdays and weekends?


## Responses
For this assignment, we will be using the dplyr package to transform data & 
the base plotting package to create the plots


```{r warning=FALSE, message=FALSE}
library(dplyr)
```


## Loading & processing the data

Unzip the file, read the contents of the file, & run a str method on the data

```{r}
a <- unzip("activity.zip")
activity <- read.csv(a, stringsAsFactors = FALSE)
activity <- tbl_df(activity)
str(activity)
```


Since the date field doesn't look like it's formatted, do it now

```{r}
activity$date <- as.Date(activity$date)
str(activity)
```


## What is the mean total number of steps taken per day?

#### Calculate the total number of steps taken per day
Ignore the missing values for this calculation

```{r, warning = FALSE, message = FALSE}
total_daily_steps <- activity %>% 
                        group_by(date) %>%
                        filter(!is.na(steps)) %>%       # ignores missing values
                        summarise(tot = sum(steps))

head(total_daily_steps,10)
```


#### Create a histogram of the total number of steps:

```{r}
hist(total_daily_steps$tot, 10, main = "Histogram of steps \n taken each day",
     xlab = "Steps", ylab = "Days")
```


#### Mean number of steps

```{r, echo = TRUE}
mean(total_daily_steps$tot, na.rm = TRUE)
```


#### Median number of steps

```{r, echo = TRUE}
median(total_daily_steps$tot, na.rm = TRUE)
```


## What is the average daily pattern?

##### Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

A: This requires the data to be aggregated by the interval. Note that the xlabels refer to 5 minute interval time slots, ranging from 00:00 hours to 24:00 hours.


```{r}
steps_per_interval <- activity %>%
                        group_by(interval) %>%
                        filter(!is.na(steps)) %>%       
                        summarise(steps_pi = mean(steps))

plot(y = steps_per_interval$steps_pi, x = steps_per_interval$interval, 
        type = "l", xlab = "5 minute intervals", main = "Pattern of Daily Activity", ylab = "Average steps")

```


#### Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

A: It appears that the most number of steps, on average, are taken at 835am (206 steps)


```{r, message=FALSE, error = FALSE, warning = FALSE, tidy = TRUE}
filter(steps_per_interval, steps_pi == max(steps_pi))

```


## Imputing missing values

##### Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)

A: There are 2304 missing values in the dataset

```{r, echo = TRUE}
sum(is.na(activity$steps))
```


#### Devise a strategy for filling in all of the missing values in the dataset. 

A: The assumption is that since people are creatures of habit, hence activities 
will follow a daily pattern.  This is the logic used to impute values where there are missing values, & hence the 5 minute interval mean is used. 



#### Create a new dataset that is equal to the original dataset but with the missing data filled in.

```{r }
steps <- data.frame(activity$steps)
steps[is.na(steps),] <- ceiling(tapply(X=activity$steps,INDEX=activity$interval,
                                       FUN=mean,na.rm=TRUE))
new <- cbind(steps, activity[,2:3])
colnames(new) <- c("Steps", "Date", "Interval")
new <- tbl_df(new)

new_total_daily_steps <- new %>%
                        group_by(Date) %>%
                        summarise(t = sum(Steps))
```


#### Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?

The blue vertical lines represent the median steps taken for each plot

```{r, , fig.width = 10, fig.height = 5}

par(mfrow = c(1, 2))

# with na values eliminated
hist(total_daily_steps$tot, 10, main = "Steps taken each day \n NA values
     removed", xlab = "Steps", ylab = "Days", ylim = c(0, 25))
abline(v = median(total_daily_steps$tot), col = 4, lwd  = 4)

# with NA values imputed
hist(new_total_daily_steps$t, 10, main = "Steps taken each day \n NA values
     imputed", xlab = "Steps", ylab = "Days", ylim = c(0, 25))
abline(v = median(new_total_daily_steps$t), col = 4, lwd  = 4)

```

#### Mean number of steps:

```{r, echo = FALSE}
mean(new_total_daily_steps$t, na.rm = TRUE)
```


#### Median number of steps

```{r, echo = FALSE}
median(new_total_daily_steps$t, na.rm = TRUE)
```

It appears that introducing missing data as an imputed value has caused both the mean & median values to increase:

```{r}
mean(new_total_daily_steps$t) - mean(total_daily_steps$tot) 

median(new_total_daily_steps$t) - median(total_daily_steps$tot)

```


## Are there differences in activity patterns between weekdays and weekends?
The dataset with the missing values filled in is used for this exercise. 


#### Create a new factor variable in the dataset with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.


```{r}
new <- mutate(new, wd = weekdays(Date))
new$fw[new$wd == "Sunday" | new$wd == "Saturday"] <- "Weekend"
new[is.na(new$fw),5] <- "Weekday"


wday <- new %>%
                group_by(Interval) %>%
                filter(fw == 'Weekday') %>%
                filter(!is.na(Steps)) %>%
                summarise(steps_wday = mean(Steps))

wend <- new %>%
                group_by(Interval) %>%
                filter(fw == 'Weekend') %>%       
                summarise(steps_wend = mean(Steps))

```



#### Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). 


```{r, fig.height= 5, fig.width = 10}

par(mfrow = c(1,2))

plot (y = wday$steps_wday, x = wday$Interval, type = "l", ylim = c(0, 250), 
      xlab = "5-minute intervals (Weekdays)", ylab = "Avg No. of Steps")
plot (y = wend$steps_wend, x = wend$Interval, type = "l", ylim = c(0, 250),
      xlab = "5-minute intervals (Weekends)", ylab = "Avg No. of Steps")

```

There's a distinct difference in activity levels between weekdays & weekends.
This random person may have had the opportunity to sleep in on the weekend (note
the later activity start in the second plot). I speculate that this person has 
a desk job (not much steps during the day), while over the weekend, they are 
more active throughout the day.